Continuous gravitational waves in the lab: recovering audio signals with a table-top optical microphone

### 2021-04-08 paper/response/AJPRefereeResponse.tex

\subsection*{Editors comments (Joseph Amato)}

This is an addendum to my decision letter and the two reviews of your manuscript.  I am satisfied that the reviewers read your work carefully, and that their recommendations are fair, thoughtful, and potentially useful.  While I sympathize with R1’s concern that the manuscript is better suited to a research journal, I think it could be revised to make a valuable contribution to AJP, which, as you know, is geared toward pedagogy.  So as I read your manuscript, I recorded things that were not sufficiently clear to me, supposing that the same things might not be clear to many AJP readers who might be very interested in duplicating your apparatus.  Here are some of my concerns.  \\

\noindent 1. Simple things: Convert footnotes to endnotes, blending them in with the bibliography.  Also, use US English (color, not colour; defense, not defence, etc.). 
\begin{itemize}
\item Done: we have moved the footnotes to the bibliography
\item \todo{Work in progress, please correct any that have not been fixed yet}
\end{itemize}

\noindent 2. Section III.  How was the Fourier transform generated?  (software?)
\begin{itemize}
\item We have added a sentence (``The Discrete Fourier transform was calculated using the NumPy package for Python.''), a citation (to the creators of NumPy), and a footnote (to the documentation).
\end{itemize}

@@

\noindent 3.Section IV.  The term “Markov process” may not be familiar to readers.  Figure 4: The three shades (or are there four?) indicating probability do not show up well in color, and worse, will be indistinguishable in black/white.  I’m guessing that’s the reason I didn’t understand why certain paths were grey and not black.  While I applaud your efforts to explain the Viterbi algorithm (VA), I still think there’s room for improvement.  I don’t understand how it improves the signal to noise ratio. 
\begin{itemize}
\item We have added a sentence to explain Markov process: ``In a Markov process, the current state depends only on the previous state (in this case the state is the frequency of the signal).''
\item \todo{image colours}
\item \han{[Hannah] The hidden Markov model and Viterbi algorithm's strength is in it's ability to find randomly wandering signals (rather than an improvement in SNR). The single constant tone of section III is recovered easily using a Fourier transform, however if the signal frequency changes randomly over time, this will reduce the Fourier transform's ability to identify the signal as it is spread across multiple frequency bins. [Added sentence at start of section IV, para 2 to explain that Fourier transform of whole data set is not well suited here]}
\end{itemize}

\noindent 4. Figure 5 is described as a heatmap.  Please explain, in the caption, what the color coding indicates.  Can you compare what is shown to what would be detected without using the VA?  Is this “heatmap” the same representation as is often shown in LIGO publications?
\begin{itemize}
\item An explanation of the color coding has been added to the caption.
\item \todo{compare to without VA}
\item \todo{compare to LIGO literature}
\end{itemize}

\noindent 5. Section V.  What do you mean by an optical microphone device?  “Cloth screen” over the photodiode?  Fig. 6 is unnecessary; the differences from Fig. 1 are explained clearly in the text.
\begin{itemize}
\item Optical microphone: we have rearranged the text so that the optical microphone concept is introduced in a separate paragraph and expanded our discission of it to improve clarity. (Section V, paragraph 2)
\item Fig. 6: we have removed this figure and moved the photo to the Supplementary Material.
\item Cloth screen: we have added an additional sentence explaining the cloth screen mount and the source of the cloth screen cover (it is the ``grill cloth'' and plastic frame from a dismantled commercial speaker).
\end{itemize}

@@

\noindent 6. Figure 7: why is there a -100 dB difference in the signal level from 7a to 7b?  
\begin{itemize}
\item \todo{...}
\end{itemize}

\noindent 7. Much of your description of the various filters should be moved to Supplementary Materials and/or referenced.  This is a physics audience.  More interesting is your observation that all filters degrade the signal in some way, i.e., they are all compromises.  You could usefully expand that argument without becoming overly mathematical.
\begin{itemize}
\item Agreed, we have moved the majority of this section to Supplementary Matrial
\item \todo{decide what we will show in main text}
\item \todo{expansion on compromises discussion}
\end{itemize}

\noindent 8.Where in your flowchart are the various filters you describe?  Are the codes for them readily available?  How does one implement a filter?  
\begin{itemize}
\item Added an explanation in the caption that the filters are performed separately, after the data is gathered.
\item Most of the code is readily available on the provided GitHub with the exception of the Weiner filtering results. The provided code is documented and commented if someone wanted to replicate the filter implementations.
\item \todo{We may need to have access to Changrong's processing code}
\item \todo{...}
\end{itemize}

@@

\noindent So you see that I’m asking for a major revision of your manuscript.  I hope this does not discourage you.  My goal is to expand the size of the audience reading your article, and to increase its utility for that group of educators.  I look forward to a revised manuscript.

\subsection*{Comments from R3}

The paper describes a table top interferometer for demonstrating the basic operating principles of gravitational wave (GW) detectors. Audio signals are used to demonstrate gravitational wave data analysis techniques for the detection and estimation of signals in noisy data. The paper is well written and will be of interest to educators who wish to provide some hands-on training to students in GW astronomy. I have some comments below that the authors should address.\\

\noindent 
-- The data analysis problems discussed in this paper are related to continuous wave searches. However, I did not find a mention of the key feature involved in these searches, namely, Doppler modulation of the signal frequency caused by the rotation of the Earth. This is the principal reason behind the extremely high computational cost of CW searches. A discussion of doppler modulation should be included. (It would be neat if the setup could be extended in the future to show the Doppler effect.)
\begin{itemize}
\item A mention has been made in the Future Work as to demonstrating the Doppler effect.
\item \todo{...}
\end{itemize}

@@

\noindent
-- Is the HMM defined properly in Sec. IV A? The observables here are the amplitudes of the spectrogram pixels at a given time index and the hidden markov process is the underlying drift of frequency and amplitude of the signal. I was expecting to see a conditional pdf that connects the state of the underlying (suitably discretized) process to the observables. I don't see it.
\begin{itemize}
\item \todo{...}
\end{itemize}

\noindent
-- Is Eq. 1 correct? The LHS is a marginal probability while the RHS is a joint probability. Shouldn't there be a marginalization on the RHS?
\begin{itemize}
\item \todo{...}
\end{itemize}

\noindent
-- The normalization used for $F(t_i, f_j)$ should be defined explicitly. 
\begin{itemize}
\item \todo{...}
\end{itemize}

\noindent
-- Looking at Fig. 3, it appears that the noise is not white and rises in power at lower frequencies. How is the color of the noise taken into account, if at all, in the HMM or the Viterbi algorithm? A discussion of this point should be included although it is fine to assume white noise as a first approximation. 
\begin{itemize}
\item Mention of this is now made both at the introduction of the plot and in the Viterbi algorithm section -- this implementation does assume white noise.
\end{itemize}

@@

\noindent
-- While the authors have applied standard filtering methods in Sec. VC, the removal of high power line features from data in GW detectors is done using more sophisticated approaches. For example, cross-channel regression, linear predictive filtering, Kalman filtering etc. There are also some non-linear and non-parametric methods that have been explored in the GW literature. This point should be mentioned in the paper.
\begin{itemize}
\item \todo{...}
\end{itemize}

\noindent
-- The axes label for frequency is not consistent across the figures: in some plots it is "f/Hz" and others "f(Hz)".
\begin{itemize}
\item \todo{...}
\end{itemize}

\noindent
-- This is not a comment but just my curiosity: How strong is the effect of mechanical vibrations of the camera or photodiode mount on the observed signal?

\begin{itemize}
\item This is unknown to us but has now been mentioned in Future Work.
\end{itemize}

### 2021-04-13 paper/response/AJPRefereeResponse.tex

\documentclass{article}

\usepackage{fullpage}
\usepackage{xcolor}
\newcommand{\todo}{\textcolor{blue}}
\newcommand{\han}{\textcolor{orange}}

\begin{document}

\todo{Opening comments}
Dear Joseph Amato,

We thank you and the referees for your helpful comments and suggestions. 
We address the comments below in turn.

@@

\noindent 3.Section IV.  The term “Markov process” may not be familiar to readers.  Figure 4: The three shades (or are there four?) indicating probability do not show up well in color, and worse, will be indistinguishable in black/white.  I’m guessing that’s the reason I didn’t understand why certain paths were grey and not black.  While I applaud your efforts to explain the Viterbi algorithm (VA), I still think there’s room for improvement.  I don’t understand how it improves the signal to noise ratio. 
\begin{itemize}
\item We have added a sentence to explain Markov process: ``In a Markov process, the current state depends only on the previous state (in this case the state is the frequency of the signal).''
\item \todo{image colours}
\item \han{[Hannah] The hidden Markov model and Viterbi algorithm's strength is in it's ability to find randomly wandering signals (rather than an improvement in SNR). The single constant tone of section III is recovered easily using a Fourier transform, however if the signal frequency changes randomly over time, this will reduce the Fourier transform's ability to identify the signal as it is spread across multiple frequency bins. [Added sentence at start of section IV, para 2 to explain that Fourier transform of whole data set is not well suited here]}
\end{itemize}

\noindent 4. Figure 5 is described as a heatmap.  Please explain, in the caption, what the color coding indicates.  Can you compare what is shown to what would be detected without using the VA?  Is this “heatmap” the same representation as is often shown in LIGO publications?
\begin{itemize}
\item An explanation of the color coding has been added to the caption.
\item \todo{compare to without VA}
\item \todo{compare to LIGO literature}
\end{itemize}

\noindent 5. Section V.  What do you mean by an optical microphone device?  “Cloth screen” over the photodiode?  Fig. 6 is unnecessary; the differences from Fig. 1 are explained clearly in the text.
\begin{itemize}
\item Optical microphone: we have rearranged the text so that the optical microphone concept is introduced in a separate paragraph and expanded our discission of it to improve clarity. (Section V, paragraph 2)
\item Fig. 6: we have removed this figure and moved the photo to the Supplementary Material.
\item The cloth screen was used to reduce the intensity of light on the photodiode and avoid saturating the analog-to-digital converter. It was indeed an improvised solution using the ``grill cloth'' and plastic frame from the dismantelled commercial speaker used in the demonstration. We have added a sentence to Section V.A. to clarify.
\end{itemize}

@@

\noindent 6. Figure 7: why is there a -100 dB difference in the signal level from 7a to 7b?  
\begin{itemize}
\item \todo{...}
\end{itemize}

\noindent 7. Much of your description of the various filters should be moved to Supplementary Materials and/or referenced.  This is a physics audience.  More interesting is your observation that all filters degrade the signal in some way, i.e., they are all compromises.  You could usefully expand that argument without becoming overly mathematical.
\begin{itemize}
\item Agreed, we have moved the majority of this section to Supplementary Matrial
\item \todo{decide what we will show in main text}
\item \todo{expansion on compromises discussion}
\end{itemize}

\noindent 8.Where in your flowchart are the various filters you describe?  Are the codes for them readily available?  How does one implement a filter?  
\begin{itemize}
\item The filters are performed separately after the data is gathered. We have added a sentence to the figure caption to clarify this. 
\item Most of the code is readily available on the provided GitHub with the exception of the Weiner filtering results. The provided code is documented and commented if someone wanted to replicate the filter implementations.
\item \todo{We may need to have access to Changrong's processing code}
\item \todo{...}
\end{itemize}

@@

The paper describes a table top interferometer for demonstrating the basic operating principles of gravitational wave (GW) detectors. Audio signals are used to demonstrate gravitational wave data analysis techniques for the detection and estimation of signals in noisy data. The paper is well written and will be of interest to educators who wish to provide some hands-on training to students in GW astronomy. I have some comments below that the authors should address.\\

\noindent 
-- The data analysis problems discussed in this paper are related to continuous wave searches. However, I did not find a mention of the key feature involved in these searches, namely, Doppler modulation of the signal frequency caused by the rotation of the Earth. This is the principal reason behind the extremely high computational cost of CW searches. A discussion of doppler modulation should be included. (It would be neat if the setup could be extended in the future to show the Doppler effect.)
\begin{itemize}
\item A mention has been made in the Future Work as to demonstrating the Doppler effect.
\item \todo{...}
\end{itemize}

\noindent
-- Is the HMM defined properly in Sec. IV A? The observables here are the amplitudes of the spectrogram pixels at a given time index and the hidden markov process is the underlying drift of frequency and amplitude of the signal. I was expecting to see a conditional pdf that connects the state of the underlying (suitably discretized) process to the observables. I don't see it.
\begin{itemize}
\item \todo{...}
\end{itemize}

\noindent
-- Is Eq. 1 correct? The LHS is a marginal probability while the RHS is a joint probability. Shouldn't there be a marginalization on the RHS?
\begin{itemize}
\item This equation was indeed incorrect and we thank you for the careful review. The equation should give the probability for a specific path, and is not a sum over all possibilities. We have corrected Eq. 1 and added clarification to the text. 
\end{itemize}

### 2021-04-13 paper/ifo-complexAudio.tex
going through more comments

In this section, we explore how this demonstration can be used to teach a selection of signal processing techniques. 
We use complex audio signals (such as music and speech) as a natural successor to the constant and wandering tones used in Sections~\ref{sec:single_tone} and~\ref{sec:viterbi_wandering} respectively.
As complex audio signals are not quasi-monochromatic, the Viterbi algorithm as applied in Section~\ref{sec:viterbi_wandering} is not directly applicable here. 
Instead, we use a hierarchy of passive filters which suppress noise, yet do not assume any specific form of the signal, unlike the Fourier-based maximum likelihood filter which is tuned to the sinusoidal signals in Section~\ref{sec:single_tone} and Appendix~\ref{app:sinusoid_likelihood}.

@@

We place an OSRAM BPW21 photodiode in reverse-bias over an LM358 op-amp which together make a photo-detector that produces a voltage that depends on the incident intensity. 
The photodiode records the interference pattern at roughly the same off-center position as the webcam in Sections~\ref{sec:single_tone} and~\ref{sec:viterbi_wandering}, again chosen arbitrarily.
The photodiode is mounted on a piece of cloth stretched over a plastic frame, with the electrical leads connected underneath the cloth.
This cloth screen mount was the ``grill cloth'' from the front of a dismantled commercial speaker, re-purposed. 
The voltage signal from the photo-detector is captured by an MCP3008 $10$-bit analog-to-digital converter (ADC) connected to a Raspberry Pi Model 3 v1.2~\cite{RaspberryPi:online} (henceforth: Pi), which provides a convenient means to record the photodiode data.
Together, the circuit samples the signal at $\sim 16\,{\rm kHz}$. Resources for using the Pi and photodiode are described in Appendix~\ref{app:circuit_diagram}.

Sampling any frequency component of the analog signal above the Nyquist frequency of $8\,{\rm kHz}$ leads to aliasing\footnote{Folding of frequencies greater than half the sampling rate.} into the detected range. We include an anti-aliasing Sallen-Key filter~\cite{sallen_key_filter} tuned to $16\,{\rm kHz}$ before the ADC to prevent this from happening.
This component attenuates any frequencies above $8\,{\rm kHz}$, before they are digitally sampled. We also place a second cloth screen over the face of the photodiode to reduce the incident intensity and avoid saturating the ADC -- an improvised, physical solution that could instead be replaced by scaling down the voltage electronically. This cloth screen was also re-purposed grill cloth from a commercial speaker.
% It’s not clear what limited the sample rate to 16kHz, but it was likely non-optimal reading of the ADC by the Pi script.
% The ADC used (MCP3008) is quoted at 200kHz (or ksps, kilo samples per second).

%An updated schematic and a photograph of the entire optical microphone are shown in the left and right panels of Fig.~\ref{fig:ifo_schematic_podo}, respectively.
See Appendix~\ref{app:circuit_diagram} for a diagram of the circuit. 
\han{[Hannah] I've removed the second schematic and photo for now - could be an option for the photo to go in the appendix / supplement if we want to show the physical set up?}

@@

@@

Here we present the results of two analyses using advanced signal processing techniques.
They are briefly described and the results shown (see Supplementary Material for details).

Firstly, we consider two signal processing techniques used in combination: the cascaded notch and the Wiener filter. 
A notch filter removes signals within a specific frequency range \han{ref}. 
In our case, we would like to remove the $50\,{\rm Hz}$ mains noise and harmonics. 
Therefore, we use a cascaded notch filter where each notch is centered at each of the harmonics. 
The Wiener filter is an advanced statistical technique that makes use of statistical information from the speech data and noise. 
It amplifies parts of the signal with high SNR while suppressing pares with low SNR.

\han{refer to figure.. } shows our results.
We find that the combined cascaded notch and Wiener filter produces a better result than either alone. 
Most of the mains noise is removed, however, the recovered voice sounds muffled and is not understandable.

Secondly, we apply a speech enhancement technique. 
Ref.~\cite{SubjectiveComparison} compares $13$ speech enhancement methods, finding the log minimum mean-square-error (logMMSE) estimator to be the best, qualitatively, at recovering speech (see also Ref.~\cite{Ephraim1984SpeechEU_logMMSE} for details on the logMMSE estimator). 
We use an existing implementation of the logMMSE~\cite{logmmse}.

\han{refer to figure} 
After applying the logMMSE estimator we see significant attenuation of the mains harmonics and general smoothing of the spectrum. 
Most of the background noise is removed, however, the logMMSE does not significantly enhance the speech as the voice sounds muffled and indistinct.
We find some improvement with music over speech. 
Simple chords and drums can be heard, however more composite sounds and complex melodies cannot be heard clearly. 
Our observations suggest that this is especially true for certain instruments, in particular flutes and violins sometimes can’t be heard at all. 
This could be a perceptual effect or a frequency dependence somewhere in the optical microphone.
Speculating, perhaps the speaker-mirror coupling is stronger at low frequencies and thus instruments like electric bass and drums are louder in the results.
To address these problems, we need to determine whether the signals that are audibly missing (the diction in the speech and complex melodies in music) are indeed being transmitted through the optical microphone at all. 
To determine this requires a better understanding of the system, as discussed in Section~\ref{sec:future_work}.

>>>>

In this section, we explore how this demonstration can be used to teach a selection of signal processing techniques. 
We use complex audio signals (such as music and speech) as a natural successor to the constant and wandering tones used in Sections~\ref{sec:single_tone} and~\ref{sec:viterbi_wandering} respectively.
As complex audio signals are not quasi-monochromatic, the Viterbi algorithm as used in Section~\ref{sec:viterbi_wandering} is not directly applicable here. 
Instead, we use a hierarchy of passive filters which suppress noise, yet do not assume any specific form of the signal, unlike the Fourier-based maximum likelihood filter which is tuned to the sinusoidal signals in Section~\ref{sec:single_tone} and Appendix~\ref{app:sinusoid_likelihood}.

@@

We place an OSRAM BPW21 photodiode in reverse-bias over an LM358 op-amp which together make a photo-detector that produces a voltage that depends on the incident intensity. 
The photodiode records the interference pattern at roughly the same off-center position as the webcam in Sections~\ref{sec:single_tone} and~\ref{sec:viterbi_wandering}, again chosen arbitrarily.
% I think this is probably not needed as I believe the editor comment is referring to the cloth screen over the photodiode, which is addressed below. 
%The photodiode is mounted on a piece of cloth stretched over a plastic frame, with the electrical leads connected underneath the cloth.
%This cloth screen mount was the ``grill cloth'' from the front of a dismantled commercial speaker, re-purposed. 
The voltage signal from the photo-detector is captured by an MCP3008 $10$-bit analog-to-digital converter (ADC) connected to a Raspberry Pi Model 3 v1.2~\cite{RaspberryPi:online} (henceforth: Pi), which provides a convenient means to record the photodiode data.
Together, the circuit samples the signal at $\sim 16\,{\rm kHz}$. 
Resources for using the Pi and photodiode, including a circuit diagram, are described in Appendix~\ref{app:circuit_diagram}.

Sampling any frequency component of the analog signal above the Nyquist frequency of $8\,{\rm kHz}$ leads to aliasing\footnote{Folding of frequencies greater than half the sampling rate.} into the detected range. We include an anti-aliasing Sallen-Key filter~\cite{sallen_key_filter} tuned to $16\,{\rm kHz}$ before the ADC to prevent this from happening.
This component attenuates any frequencies above $8\,{\rm kHz}$, before they are digitally sampled. We also place a cloth screen over the face of the photodiode to reduce the incident intensity and avoid saturating the ADC -- an improvised, physical solution that could instead be replaced by scaling down the voltage electronically. This cloth screen was re-purposed grill cloth from a commercial speaker.
% It’s not clear what limited the sample rate to 16kHz, but it was likely non-optimal reading of the ADC by the Pi script.
% The ADC used (MCP3008) is quoted at 200kHz (or ksps, kilo samples per second).

%An updated schematic and a photograph of the entire optical microphone are shown in the left and right panels of Fig.~\ref{fig:ifo_schematic_podo}, respectively.
%See Appendix~\ref{app:circuit_diagram} for a diagram of the circuit.

@@

The raw output of the optical microphone (with anti-aliasing) is noisy with a loud bass hum throughout. 
This can be explained by looking at the power spectral density (PSD) of the background noise (i.e., with the speaker switched off), \fixme{as shown in the top panel} of Fig.~\ref{fig:psd_noise}. 
The spectrum is dominated by AC mains noise with power from the fundamental $50\,{\rm Hz}$ mains signal up to and beyond $8$th harmonic thereof. 
The mains signal is also present (but far weaker) in the background spectrum taken with the photodiode in darkness, suggesting that ambient lighting has a large contribution. 
Besides lighting, other possible contributions to the mains signal include air conditioning and the photodiode circuit itself. 
The appearance of harmonics of the mains noise is likely due to the non-linearity discussed in Appendix~\ref{app:intensity_derivation} (see also Ref.~\cite{feynman}). 
The spectrum in Fig.~\ref{fig:psd_noise} also has a broad feature at around $750\,{\rm Hz}$, the origin of which is yet to be determined.

@@

In this section we present results of two advanced signal processing analysis techniques applied to the optical microphone recordings. 
The techniques are only briefly described here and we refer the reader to the accompanying Supplementary Material for further details and some other analysis techniques.

Firstly, we consider two signal processing techniques used in combination: the cascaded notch and the Wiener filter. 
A notch filter removes signals within a specific frequency range \han{ref}. 
In this case, our aim is to remove the $50\,{\rm Hz}$ mains noise and harmonics, therefore we use a cascaded notch filter where each notch is centered at each of the harmonics. 
The Wiener filter is an advanced statistical technique that makes use of statistical information from the speech data and noise. \han{ref}
It amplifies parts of the signal with high SNR while suppressing pares with low SNR. 
The results of the combined cascaded notch and Wiener filter are shown in the third row of Fig.~\fixme{XXXXX} where the timeseries and spectrum are shown by the left and right panels respectively. 
Most of the mains noise is removed, however, the recovered voice sounds muffled and is not understandable.

Secondly, we apply a speech enhancement technique. 
Ref.~\cite{SubjectiveComparison} compares $13$ speech enhancement methods, finding the log minimum mean-square-error (logMMSE) estimator to be the best, qualitatively, at recovering speech (see also Ref.~\cite{Ephraim1984SpeechEU_logMMSE} for details on the logMMSE estimator). 
We use an existing implementation of the logMMSE from~\cite{logmmse}.  
The logMMSE estimator results are shown in the bottom panel of Fig.~\fixme{XXXXX}. 
We see significant attenuation of the mains harmonics and general smoothing of the spectrum. 
Most of the background noise is removed, however, the logMMSE does not significantly enhance the speech as the voice sounds muffled and indistinct.
We find some improvement with music over speech. 
Simple chords and drums can be heard, however more composite sounds and complex melodies cannot be heard clearly. 
Our observations suggest that this is especially true for certain instruments, in particular flutes and violins sometimes can’t be heard at all. 
This could be a perceptual effect or a frequency dependence somewhere in the optical microphone.
Speculating, perhaps the speaker-mirror coupling is stronger at low frequencies and thus instruments like electric bass and drums are louder in the results.
To address these problems, we need to determine whether the signals that are audibly missing (the diction in the speech and complex melodies in music) are indeed being transmitted through the optical microphone at all. 
To determine this requires a better understanding of the system, as discussed in Section~\ref{sec:future_work}.
